Spatial and temporal decay of SARS-CoV-2 environmental contamination of the healthcare environment

### 2021-09-28 manuscript.Rmd
Fix typos.

* __Results__: The probability of detecting SARS-CoV-2 RNA in a patietn room did not vary with distance. However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)). Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)). The probability of surface elevated contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.

@@

Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the hospital and other healthcare settings. In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previoulsy uninfected patients developed COVID-19. Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].

@@

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did predict different levels of contamination with highest probability among floors followed be elevated high-touch surfaces. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020]. Alternatively, it is possible that the quality of cleaning declined as the study progressed, or that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

We found that increased clinical severity of COVID-19, measured by oxygen requirement, was associated with increased contamination of the healthcare environment. Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory procedures. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination. While we can only speculate on the mechanism for greater detection in these patients, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room. This finding matches previous description of high probability of floor contamination relative to other surfaces[@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets or aeresols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

Our study has several limitations. First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. Secondly, we did not measure differences in cleaning methods across different study units. However, cleaning practices did not undergo systematic changes during the study period and sampling was performed routinely prior to daily cleaning. Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.

>>>>

* __Results__: The probability of detecting SARS-CoV-2 RNA in a patient room did not vary with distance. However, we found that surface type predicted probability of detection, with floors and high-touch surfaces having the highest probability of detection (floors odds ratio (OR) `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_site_categoryFloor") %>% pull(Q97.5_exp) %>% signif(.,3)`); high-touch elevated OR `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "Distance Fixed" & param == "b_high_touchTRUE") %>% pull(Q97.5_exp) %>% signif(.,3)`)). Patients requiring high-flow oxygen, positive airway pressure, or mechanical ventilation were associated with increased surface contamination (OR `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Estimate_exp) %>% signif(.,3)` (95% CrI `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q2.5_exp) %>% signif(.,3)` to `r m_combined_summary %>% filter(model_name == "COVID-19 Severity") %>% filter(grepl("SignificantOxygen",param)) %>% pull(Q97.5_exp) %>% signif(.,3)`)). The probability of surface elevated contamination decayed with prolonged hospitalization, but the probability of floor detection increased with duration of the local pandemic wave.

@@

Despite infection control measures, transmission of SARS-CoV-2 infection has been observed in the hospital and other healthcare settings. In a recent study of SARS-CoV-2-infected and -uninfected patients inadvertently roomed together, 39% of exposed, previously uninfected patients developed COVID-19. Risk of SARS-CoV-2 transmission was associated with source patients who had low cycle threshold (Ct), i.e., high viral load, in their diagnostic testing [@Karan2021-cm].

@@

We sought to define the impact of time and distance on SARS-CoV-2 contamination of the healthcare environment through longitudinal sampling of surfaces within COVID-19 patient rooms. We found that distance from the patient’s bed did not inform the probability of detection of SARS-CoV-2 RNA, but the category of surface (elevated or floor, high-touch or low-touch) did. The highest probability of contamination was found with floors, followed by elevated high-touch surfaces. We also found a decrease in probability of detection among both high and low-touch surfaces with time from COVID-19 diagnosis, consistent with previous reports [@Chia2020]. While the overall probability of SARS-CoV-2 RNA detection decreased as time from diagnosis increased, the probability of SARS-CoV-2 RNA detection on floors increased with time from the onset of the local COVID-19 case wave. The unique behavior of floor surfaces raises the possibility of accumulation as SARS-CoV-2 RNA carried on the feet of healthcare personnel (HCP) or the mops of EVS teams increases with the COVID-19 patient census. This possibility is supported by other research where floor contamination was discovered to spread beyond clinical areas[@Guo2020]. Alternatively, it is possible that the quality of cleaning declined as the study progressed, or that viral RNA persists on surfaces even after terminal cleaning, leading to an accumulation effect.

We found that increased clinical severity of COVID-19, measured by oxygen requirement, was associated with increased contamination of the healthcare environment. Specifically, those subjects who required high-flow nasal cannula, continuous or bilevel positive airway pressure, or mechanical ventilation, were found to have increased odds of detecting surface SARS-CoV-2 RNA, compared to those with no oxygen requirement. This finding may be the result of either the higher viral burden of occupying patients, or the role of increased respiratory droplet and aerosol production associated with respiratory support. In a prior study, disease severity was not correlated with cycle threshold from anatomic specimens [@Abdulrahman2021], potentially supporting the contribution of respiratory care in surface contamination. In either case, our findings support CDC guidance for prolonged isolation in critically ill patients[@CDC2021].

Our study design and hypotheses were informed by our prior work with bacterial deposition within patient rooms, where some taxa have decreased probability of detection as distance changes within patient room[@Kelly2021]. Contrary to our findings with multidrug-resistant bacteria, SARS-CoV-2 detection did not vary significantly within any surface group as distance from the patient increased. The difference between the findings from our two studies is most likely related to two factors: (1) the relatively small size of patient rooms and (2) the physical spread of SARS-CoV-2 RNA through touch. We observed large differences in the probability of detecting surface contamination by surface type. Floors had high probability of contamination, which did not vary throughout the room. This is likely related to both the role of patient and HCP foot traffic in addition to the physical action of mopping which may spread body fluids containing viral RNA within the patient’s room. Our findings match previous description of high probability of floor contamination relative to other surfaces[@Chia2020]. In contrast to floors, walls had significantly lower probability of contamination. This may be due to the decreased settling of respiratory droplets or aerosols on vertical surfaces, or the decreased role of touch or secondary contamination through cleaning.

Our study has several limitations. First, we used RT-qPCR for the detection of SARS-CoV-2 RNA and did not use viral culture to determine the viability from surfaces. Second, we did not measure differences in cleaning methods across different study units. However, cleaning practices did not undergo systematic changes during the study period, and sampling was performed routinely prior to daily cleaning. Finally, we were not able to measure SARS-CoV-2 viral loads directly from human subjects, which would have permitted us to better explore the contribution of differences in shedding between subjects.
